biological data preperation

### 2019-01-14 src/biological_data_prep.Rmd
moving through presence point check - rerun loop just in case

>>>>

This is interesting...
chl and o2 are missing the same number of points - they come from the biomer dataset; salinity and temp come from glorys and have different values.

I wonder how many of these points represent the same lat-long-depth... need to subset where chl_depth OR o2_depth OR salinity_depth OR temp_depth == NA These columns are not together

```{r}
presence_wdepth2 <-  presence_wdepth[!complete.cases(presence_wdepth[28:37]), ]
nrow(presence_wdepth2) #3349
```
ok so there are 3349 points without any data.let's map these

```{r map presences with depth without any env. data at depth}
map <- getMap(resolution = "low") #creates an object called map at low resoultion
plot(map, xlim = c(-70, -43), ylim =c(38,70), asp = 1) #the x and y lim are the long-lat bounds of the map
points(presence_wdepth2$decimalLongitude, presence_wdepth2$decimalLatitude, col = "red") #this adds points to the mapet", xlab = "Longitude", ylab = "Latitude")
```
hmm everywhere...

Is there a particular time-slice that the values appear in?

```{r}
npresence_wdepth2mmyy <- with(presence_wdepth2, table(year, month))
npresence_wdepth2mmyy
```

hmm there is a lot... how does this compare to the original dataset

```{r}
data_aeammyy <- with(data_aea, table(year, month))
data_aeammyy
```

ok so 2014_07 and 1999_07 are both missing all data. Re-run the loop for these two time-periods (just add the nectcdfs for these files in renetcdf folder).

First you need to put the observations (data_aea) into a spatialpointsdataframe.

```{r}
fmatch("longitude_meters", names(data_aea)) #19
fmatch("latitude_meters", names(data_aea)) #20
```

```{r}
library(sp)
xy <- data_aea[ ,c(19,20)] # This is to tell R where the coordinates are (in column 19 and 20). Note that the column order needs to be longitude, latitude
data_aea_sp <- SpatialPointsDataFrame(coords = xy, data = data_aea, proj4string = CRS("+proj=aea +lat_1=50 +lat_2=70 +lat_0=40 +lon_0=-60 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs")) # The CRS is used here is for the albers equal area projection.
```

```{r}
netcdf_list <- list.files("../data/env/renetcdf", pattern = '*.nc', full.names = TRUE) #true means the full path is included
no_netcdf <- length(netcdf_list) #for the loop - need to know how many files to cycle through
netcdf_name <- list.files("../data/env/renetcdf", pattern = '*.nc', full.names = FALSE) #false means the path is not includedCores
library (ncdf4)
library(sp)
library(rgdal)
for (i in 1:no_netcdf) {  
  library(raster)
  print(netcdf_name[i]) #this just prints the name of the netCDF R is working one
  brkyr <- as.integer(sapply(strsplit(netcdf_name[i], "_"), "[[", 1)) # extracting the first part of the netcdf filename (which is the year)
  brkmth <- as.integer(sapply(strsplit(netcdf_name[i], "_"), "[[", 2)) # extracting the second part of the netcdf filename (which is the month)
  brkvar <- (sapply(strsplit(netcdf_name[i], "_"), "[[", 3)) # extracting the third part of the netcdf (inc.nc)
  temp_brick <- brick(netcdf_list[i], lvar = 4)
  temp_brick <- projectRaster(temp_brick, aea) 
    for (j in 1:nrow(data_aea_sp)) {  
      de <- data_aea_sp$depthlayerno[[j]]  # a variable for the observation depth layer
      yr <- (data_aea_sp$year[j])  # a variable for the observation year
      mth <- (data_aea_sp$month[j])  # a variable for the observation month
          if (brkyr == yr & brkmth == mth & brkvar == "temp.nc"){
              if (is.na(de)){
                data_aea_sp$temp_depth[j] <- NA
              } else  
                data_aea_sp$temp_depth[j] <- extract(x=temp_brick[[de]], y = data_aea_sp[j, ])
          } else if (brkyr == yr & brkmth == mth & brkvar == "salinity.nc") {
              if (is.na(de)){
                data_aea_sp$salinity_depth[j] <- NA
              } else  
                data_aea_sp$salinity_depth[j] <- extract(x=temp_brick[[de]], y = data_aea_sp[j, ]) 
          } else if (brkyr == yr & brkmth == mth & brkvar == "chl.nc") {
              if (is.na(de)){
                data_aea_sp$chl_depth[j] <- NA
              } else  
                data_aea_sp$chl_depth[j] <- extract(x=temp_brick[[de]], y = data_aea_sp[j, ]) 
          } else if (brkyr == yr & brkmth == mth & brkvar == "o2.nc") {
              if (is.na(de)){
                data_aea_sp$o2_depth[j] <- NA
              }
          }

    }
}
testdata_aea <- as.data.frame(data_aea_sp)

lets see if that made any difference...

```{r}
presence_wdepthv2 = data_aea[!is.na(data_aea$depth), ]
nrow(presence_wdepthv2) #10178 rows - same as before
head(presence_wdepthv2)
```

```{r}
npresence_wdepth2mmyy2 <- with(presence_wdepthv2, table(year, month))
npresence_wdepth2mmyy2
```

no change...

ok is there a depth layer in particular that is an issue?
maybe some are missing a depthlayerno?

```{r}
presence_wdepth_nolayerno <- presence_wdepth2[is.na(presence_wdepth2$depthlayerno), ]
nrow(presence_wdepth_nolayerno) #0 
head(presence_wdepth_nolayerno)
```
ok so all have a depthlayerno attached

check for count of each depth layer

```{r}
whichdepths <- unique(presence_wdepth2$depthlayerno) #makes a variable of institutions
whichdepthsordered <- sort(whichdepths) #just puts the list in 
whichdepthsordered
```
quite a wide group... look at frequency

```{r count depthlayerno for data with depth but missing values}
whichdepthscount <- count(presence_wdepth2, "depthlayerno")
whichdepthscount
```
very variable...

How many rows are missing all depth data?

```{r}
presence_wdepthonly <- subset(presence_wdepth2, select = -c(mlp_surface, ssh_surface, temp_surface, salinity_surface, o2_surface, chl_surface))
head(presence_wdepthonly)
```

get the col numbers for the depth data
```{r}
fmatch("chl_depth", names(presence_wdepthonly)) #28
fmatch("o2_depth", names(presence_wdepthonly)) #29
fmatch("salinity_depth", names(presence_wdepthonly)) #30
fmatch("temp_depth", names(presence_wdepthonly)) #31
```

now subset rows where all depth data is missing
```{r}
presence_wdepthonly_allmissing <-  presence_wdepthonly[(is.na(presence_wdepthonly$temp_depth)) & (is.na(presence_wdepthonly$o2_depth)) & (is.na(presence_wdepthonly$salinity_depth)) & (is.na(presence_wdepthonly$chl_depth)), ]
nrow(presence_wdepthonly_allmissing) #2178
head(presence_wdepthonly_allmissing)
```

so there are 2178 out of 3349 observations that are missing all data for depth. so what about the remaining 1171 points?

want to remove from the observations in presence_wdepthonly (all 3349 observations) the ones that appear in presence_wdepthonly_allmissing (2178 observations)

some code from [stack exchange](https://stackoverflow.com/questions/10907359/r-remove-rows-from-one-data-frame-that-are-in-another)

```{r}
presence_wdepthonly_somemissing <-  rbind(presence_wdepthonly_allmissing, presence_wdepthonly)
presence_wdepthonly_somemissing <- presence_wdepthonly_somemissing[!duplicated(presence_wdepthonly_somemissing, fromLast = FALSE) & !duplicated(presence_wdepthonly_somemissing, fromLast = TRUE), ]
nrow(presence_wdepthonly_somemissing)
head(presence_wdepthonly_somemissing)
```

ok before going further, just rerun the loop again on all the data...

First you need to put the observations (data_aea) into a spatialpointsdataframe.
```{r}
library(sp)
xy <- data_aea[ ,c(19,20)] # This is to tell R where the coordinates are (in column 18 and 19). Note that the column order needs to be longitude, latitude
data_aea_sp <- SpatialPointsDataFrame(coords = xy, data = data_aea, proj4string = CRS("+proj=aea +lat_1=50 +lat_2=70 +lat_0=40 +lon_0=-60 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs")) # The CRS is used here is for the albers equal area projection.
```

```{r}
netcdf_list <- list.files("../data/env/netcdf", pattern = '*.nc', full.names = TRUE) #true means the full path is included
no_netcdf <- length(netcdf_list) #for the loop - need to know how many files to cycle through
netcdf_name <- list.files("../data/env/netcdf", pattern = '*.nc', full.names = FALSE) #false means the path is not includedCores
library (ncdf4)
library(sp)
library(rgdal)
for (i in 1:no_netcdf) {  
  library(raster)
  print(netcdf_name[i]) #this just prints the name of the netCDF R is working one
  brkyr <- as.integer(sapply(strsplit(netcdf_name[i], "_"), "[[", 1)) # extracting the first part of the netcdf filename (which is the year)
  brkmth <- as.integer(sapply(strsplit(netcdf_name[i], "_"), "[[", 2)) # extracting the second part of the netcdf filename (which is the month)
  brkvar <- (sapply(strsplit(netcdf_name[i], "_"), "[[", 3)) # extracting the third part of the netcdf (inc.nc)
  temp_brick <- brick(netcdf_list[i], lvar = 4)
  temp_brick <- projectRaster(temp_brick, aea) 
    for (j in 1:nrow(data_aea_sp)) {  
      de <- data_aea_sp$depthlayerno[[j]]  # a variable for the observation depth layer
      yr <- (data_aea_sp$year[j])  # a variable for the observation year
      mth <- (data_aea_sp$month[j])  # a variable for the observation month
          if (brkyr == yr & brkmth == mth & brkvar == "temp.nc"){
              data_aea_sp$temp_surface[j] <- extract(x=temp_brick[[1]], y = data_aea_sp[j, ]) 
              if (is.na(de)){
                data_aea_sp$temp_depth[j] <- NA
              } else  
                data_aea_sp$temp_depth[j] <- extract(x=temp_brick[[de]], y = data_aea_sp[j, ])
          } else if (brkyr == yr & brkmth == mth & brkvar == "salinity.nc") {
              data_aea_sp$salinity_surface[j] <- extract(x=temp_brick[[1]], y = data_aea_sp[j, ]) 
              if (is.na(de)){
                data_aea_sp$salinity_depth[j] <- NA
              } else  
                data_aea_sp$salinity_depth[j] <- extract(x=temp_brick[[de]], y = data_aea_sp[j, ]) 
          } else if (brkyr == yr & brkmth == mth & brkvar == "chl.nc") {
              data_aea_sp$chl_surface[j] <- extract(x=temp_brick[[1]], y = data_aea_sp[j, ]) 
              if (is.na(de)){
                data_aea_sp$chl_depth[j] <- NA
              } else  
                data_aea_sp$chl_depth[j] <- extract(x=temp_brick[[de]], y = data_aea_sp[j, ]) 
          } else if (brkyr == yr & brkmth == mth & brkvar == "o2.nc") {
              data_aea_sp$o2_surface[j] <- extract(x=temp_brick[[1]], y = data_aea_sp[j, ]) 
              if (is.na(de)){
                data_aea_sp$o2_depth[j] <- NA
              } else  
                data_aea_sp$o2_depth[j] <- extract(x=temp_brick[[de]], y = data_aea_sp[j, ]) 
          } else if (brkyr == yr & brkmth == mth & brkvar == "mlp.nc") {
              data_aea_sp$mlp_surface[j] <- extract(x=temp_brick[[1]], y = data_aea_sp[j, ])
          } else if (brkyr == yr & brkmth == mth & brkvar == "ssh.nc") {
              data_aea_sp$ssh_surface[j] <- extract(x=temp_brick[[1]], y = data_aea_sp[j, ])

          }

    }
}
data_aea <- as.data.frame(data_aea_sp)
write.csv(data_aea,"../output/bio/presence_points_without_envdata_relooped.csv", row.names = FALSE)
head(data_aea_sp$chl_depth)
```
note that if you wanted to just loop through the netcdfs and create a ton of raster bricks with the same name as the netcdf, the following code you wrote before works

netcdf_list <- list.files("../data/env/netcdf", pattern = '*.nc', full.names = TRUE) #true means the full path is included
no_netcdf <- length(netcdf_list) #for the loop - need to know how many files to cycle through
netcdf_name <- list.files("../data/env/netcdf", pattern = '*.nc', full.names = FALSE) #false means the path is not included
library (ncdf4)
for (i in 1:no_netcdf) {
  print(netcdf_name[i]) #this just prings the name of the netCDF R is working on
  temp_brick <- brick(netcdf_list[i], lvar = 4)
  temp_brick <- projectRaster(temp_brick, aea) #aea is an existing raster in the projection I want that I created (environmental_data_preperation)
  assign(netcdf_name[i], temp_brick) #this asigns the netCDF name to the raster brick
}

```{r}

change dataset back into a dataframe and then save as a .csv

```{r}
